Emergent dual scaling of riverine biodiversity

### 2021-03-23 manuscript_pnas.Rmd
edits: maintext, minor

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show that ecosystem size and complexity play equally-important roles in scaling biodiversity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, suggesting the general power-law scaling along the two orthogonal dimensions of ecosystem structure, i.e., size and complexity. In support of this theoretical prediction, the robust dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.

**Significance** Larger ecosystems support more species; this ubiquitous pattern is the foundation of current conservation schemes. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such complexity in controlling biodiversity is largely unknown. Here, we use theory and extensive fish community data to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparable to the size effect, and this pattern has emerged regardless of ecological contexts. The dual scaling of biodiversity may be a pervasive feature that has far-reaching implications for biodiversity conservation.

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity controls biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

@@

While the prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

>>>>

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show that ecosystem size and complexity play equally-important roles in scaling biodiversity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, indicating a general power-law scaling along the two orthogonal dimensions of ecosystem structure, i.e., size and complexity. In support of this theoretical prediction, the robust dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.

**Significance** Larger ecosystems support more species; this ubiquitous pattern is the foundation of current conservation schemes. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such complexity in scaling biodiversity patterns is largely unknown. Here, we use theory and extensive fish community data to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparable to the size effect, and this pattern has emerged regardless of ecological contexts. The dual scaling of biodiversity may be a pervasive feature that has far-reaching implications for biodiversity conservation.

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

@@

While the prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.
